An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument

### 2016-01-12 Papers/InProgress/IJCAI/paper.tex
paper improvements

\section{Introduction}\label{sec:introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticise, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{Sobel} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported their successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
their work has received a media repercussion on a global scale\footnote{A
  small collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

The in-depth computational analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similar to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL) even though G\"odel's proof is actually
formulated in \emph{modal} higher-order logic. To bridge between the two
logics we utilise and further improve the logic embedding
approach \cite{J23,C40}, which has already been employed succesfully in preceding related work.

@@

For more than four decades, the serious consequences of G\"odel's mistake remained unnoticed, despite intense activity focused on criticizing his argument. Especially since the discovery by Sobel \cite{Sobel} that a modal collapse is entailed by G\"odel's (or also Scott's) axioms, several variants have been proposed \cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal} attempting to avoid the modal collapse. Many of these variants omit the crucial conjunct in the definition of essence as well\footnote{As these variants also change other axioms, on which the inconsistency of G\"odel's axioms depends, it is not necessarily the case that these variants are also inconsistent.}. Opponents of the argument (e.g. Oppy \cite[p.226/227]{oppy96:_goedel_ontol_argum} \cite[p.364]{oppy00:_respon_gettin} \cite[p.1068]{oppy08:_higher_order_ontol_argum}) have also proposed parodies and other criticisms, referring to variants where the conjunct is omitted.

@@

\subsection{Improved Embedding \SFiveU}\label{sec:improvedembedding}

@@

It is easy to see that \SFiveU is at least as strong as \SFive:
universality entails that $r$ is an equivalence relation. \SFiveU is,
in fact, \emph{strictly stronger} than \SFive. \SFiveU only admits
complete\footnote{A graph is \emph{complete} iff there is a directed
  edge connecting every ordered pair of vertices.} frames, whereas
\SFive admits non-complete frames as long as all their components are
complete. In other words, in \SFiveU we face one single equivalence
class of possible worlds, while in \SFive we may deal with several
disconnected equivalence classes. For this reason, \SFiveU considered
as metaphysically more appropriate by many philosophers;
cf. \cite[p.~127]{williamson13}.

@@

The new embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QML_S5}.
\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5}
\end{figure}
%
\marginpar{@Bruno: I commented you paragraph here, since I am not
  convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
  and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5}.
With the embedding used by Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40}, the
provers still had to be given the intermediate theorem T2 and the corollary
C in order to manage to prove T3.

@@

In the embedding presented here, in Fig.~\ref{QML_S5}, advanced
syntax-sugaring effects provided by Isabelle were used to prevent
issues as enumerated above. The possibility to define boldface connectives allows us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{QML_S5} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{Scott_S5}. The embedding technique is now completely transparent to the user.

@@

Interestingly, the refutation automatically generated by \textsc{Leo-II} uses a symmetric accessibility relation, and thus requires the modal logic KB. The informal refutation described above, on the other hand, requires only the weaker modal logic K.

@@

To verify the correctness of the informal argument explained above, it was reconstructed in Isabelle/HOL, using Metis to automate the inessential parts. The essential use of the Empty Essence Lemma, on the other hand, is explicitly stated, to ensure that Isabelle is reconstructing the same argument. In fact, without the help of this lemma, Metis is not strong enough to refute G\"odel's axioms. \marginpar{ToDo: check this claim about Metis.}

\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Refutation of G\"odel's Axioms in Isabelle/HOl..} \label{InconsistencyIsabelleK}
\end{figure}

@@

\section{Conclusion}\label{sec:conclusion}

Good improvement (technical sense), Sledgehammer can still be improved
(\textsc{Leo-II} finds inconsistency when call directly in THF Syntax, but not
from within Isabelle). Without independent experiments directly in THF 
the inconsistency would thus not have been detected.

As we show in this paper using this notion of essence
leads to an inconsistent axiom system, which in a classical setting of
course deserves no further attention, since from an inconsistent
system everything can be concluded, including God's existence and
non-existence simultaneously. In other words, these literature
contributions discuss variants of the argument which from a classical
logical perspective do not deserve any further attention.

We also point to several technical challenges which, at least for the
moment, still restrain non-expert users of these systems from
independently conducting similar experiments in computational
metaphysics as reported here and in previous work.
% Technical issues detected within Isabelle-Sledgehammer; room for improvements

@@

\vfill
 bla

>>>>

\section{Introduction}\label{sec:introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticise, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{sobel2004logic} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported their successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
their work has received a media repercussion on a global scale\footnote{A
  small collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

The in-depth computational analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similar to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL) even though G\"odel's proof is actually
formulated in \emph{modal} higher-order logic (HOML). To bridge between the two
logics we utilise and further improve the logic embedding
approach \cite{J23,C40}, which has already been employed succesfully in preceding related work.

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity focused on
criticizing his argument. Especially since the discovery by Sobel
\cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent.}. Opponents of the argument
(e.g. Oppy \cite[p.226/227]{oppy96:_goedel_ontol_argum}
\cite[p.364]{oppy00:_respon_gettin}
\cite[p.1068]{oppy08:_higher_order_ontol_argum}) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

\subsection{Improved Embedding \SFiveU}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5}
\end{figure}

@@

It is easy to see that \SFiveU is at least as strong as \SFive:
universality entails that $r$ is an equivalence relation. \SFiveU is,
in fact, \emph{strictly stronger} than \SFive. \SFiveU only admits
complete\footnote{A graph is \emph{complete} iff there is a directed
  edge connecting every ordered pair of vertices.} frames, whereas
\SFive admits non-complete frames as long as all their components are
complete. In other words, in \SFiveU we face one single equivalence
class of possible worlds, while in \SFive we may deal with several
disconnected equivalence classes. In fact, for this reason, \SFiveU is
considered as metaphysically more appropriate by some philosophers;
cf. \cite[p.~127]{williamson13}.

@@

The new embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QML_S5}.
%
\marginpar{@Bruno: I commented you paragraph here, since I am not
  convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
  and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5}.
With the embedding used by Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40}, the
provers still had to be given the intermediate theorem T2 and the corollary
C in order to manage to prove T3.

@@

In the embedding presented here, in Fig.~\ref{QML_S5}, advanced
syntax-sugaring effects provided by Isabelle were used to prevent
issues as enumerated above. The possibility to define boldface connectives allows us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{Scott_S5} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

@@

Interestingly, the refutation automatically generated by
\textsc{Leo-II} uses a symmetric accessibility relation, and thus
requires the modal logic KB. The informal, human-constructed
refutation described above, on the other hand, requires only the
weaker modal logic K. In our experiments \textsc{Leo-II} (like all
other HOL provers) was still to weak to automatically prove the
inconsistency already in logic K. Hence, this remains an open problem for automated
theorem provers.

@@

To verify the correctness of the informal argument explained above, it was reconstructed in Isabelle/HOL, using Metis to automate the inessential parts. The essential use of the Empty Essence Lemma, on the other hand, is explicitly stated, to ensure that Isabelle is reconstructing the same argument. In fact, without the help of this lemma, Metis is not strong enough to refute G\"odel's axioms. \marginpar{ToDo: check this claim about Metis.}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Refutation of G\"odel's Axioms in Isabelle/HOl..} \label{InconsistencyIsabelleK}
\end{figure}

\section{Conclusion}\label{sec:conclusion}
The inconsistency of G\"odel's original axioms in his ontological
argument has been detected automatically by the prover
\textsc{Leo-II}. Here we have presented a rational reconstruction and
verification of the inconsistency proof in Isabelle/HOL. Interestingly,
the inconsistency argument applies to all normal modal logics
including K.

@@

Moreover, we have reported on several technical improvements regarding
the semantic embedding approach. In fact, we have a achieved a nearly
perfect match between pen and paper presentations in HOML and the
syntax in Isabelle/HOL. As a result, the embedding of HOML in
HOL is now fully transparent and, as we think, ready for wider adoption by
philosophers.

On the other hand, there is still room for improvements in
Isabelle/HOL. For example, Isabelle/HOL's Sledgehammer tool, which
realises calls to external automated theorem provers, such as
\textsc{Leo-II}, still fails to detect the inconsistency of G\"odel's
axioms, while a direct modeling of the problem in TPTP THF syntax in
combination with a direct call of \textsc{Leo-II} succeeds. In other
words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected. There are further
technical issues regarding Sledgehammer that motivate
improvements. For example, in default setting, Sledgehammer does not
inform the user early that a proof has been found and instead silently
applies first a series of time-consuming analysis tools (e.g. it's
minimizer), before it eventually informs the user of success. For the
involved proof problems presented here (e.g. T3), this phase of
silence may take several minutes (during which the user might actually
give up on the proof attempt), while \textsc{Leo-II} already reported
success to Sledgehammer much earlier (e.g. after 2,5 seconds for T3).

Moreover, our work points to a challenge for proof transformation:
even if \textsc{Leo-II}'s proof could be automatically verified
in Isabelle, this would clearly not entail that Isabelle/HOL could
also extract from this verification a human-intuitive argument, for
example, an argument as has been presented here. For philosophers,
however, the reconstruction of such human-intuitive arguments from
machine found proofs will likely be crucial for accepting any novel results.

% Good improvement (technical sense), Sledgehammer can still be improved
% (\textsc{Leo-II} finds inconsistency when call directly in THF Syntax, but not
% from within Isabelle).

% We also point to several technical challenges which, at least for the
% moment, still restrain non-expert users of these systems from
% independently conducting similar experiments in computational
% metaphysics as reported here and in previous work.

% Technical issues detected within Isabelle-Sledgehammer; room for improvements

@@
